Historical trends in the contribution of six major causes of death to the black-white

### 2017-05-15 Manuscript/Manuscript_COD_Trends.Rmd
SH revisions

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref MMWR]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends].

The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat]. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data]. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census. Furthermore, the agreement between self-reported ethinicity on the Census  and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

@@

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@preston; @selvin_2004]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimate from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify divergent states, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.

@@

**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**eFigures 1 and 2**). However, white men underwent relatively larger mortality decreases than black men, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively greater than the general pattern, while Massachusetts and New York exhibit persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among men, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white men and in cancer's contribution to the life expectancy gap (**Figure 5**). For women, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease has been increasing over time for men in most states, with DC as a notable exception (**eFigure 5**). Recently, there has been a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with other causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of criminal assault resulted in heightened mortality from injuries among black men during the 1990s (**eFigure 7**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**eFigure 8**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white women demonstrating sharper increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Communicable disease is the rarest cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**). This spike was greater among men, especially blacks, and within states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

To investigate how much of the gap in all other causes occurred during the first year of life, we conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred *during the first year of life* to the life expectancy gap (**Figures 15 and 16**). The motivation for this analysis is that the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inequities in infant mortality. We found that 43% of the contribution of all other causes to the gap was due to differences during the first year of life in 1969. In 2013, 53% of these deaths occurred during the first year of life. Thus, the trend in the all other causes contribution partially reflects a decreasing disparity in infant mortality.

@@

Overall, the racial disparity in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century. [add: Lemann, 1991] Trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.  On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us]. On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends. [add: Case and Deaton, 2017]

Consistent with what has been reported at the national level, we found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends]. Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of equity and points to the importance of viewing the most recent estimates in light of their earlier values.

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.  States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state]. These presumably affect health and longevity, and can have especially important impacts on racial disparities when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies]. The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.  Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[add: Strumpf FE Chapter in MSE]

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated disparity contributions for a particular cause were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two]. The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum, creating interactions between cells. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period [@laveist2011black].  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic]. We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref]. In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall. As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient [@brazil2015spatial].

Secondly, our analysis does not account for immigration by the foriegn-born, or migration of native-born individuals between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC [add: US Census bureau, 2015].

@@

Racial disparities have been a fixture of American life for centuries, but their continual evolution is a topic of great relevance for public policy and social justice.  The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century. This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death. Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.

>>>>

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref MMWR]. While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied [@fuchs2016black].

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among women, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years [@harper2014trends]. However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat]. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data]. The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**).

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map]. We coulnd not consider Hispanic origin to identify race according to Hispanic ethnicity, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census. *Furthermore, the agreement between self-reported ethnicity on the Census and secondary-reported ethinicity on the death certificate has been shown to be poorer during the earlier time period and differential according to geography, which could induce measurement error into mortality and disparity estimates [ref cdc_validity_race].*

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. First, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9 [ref NCHS or NCI page on suppression], we needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

We used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-year-sex-race-age-cause category as a function of the at-risk population. The mortality rates over time within each state-year-sex-race-age-cause category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied [@preston; @selvin_2004]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.

@@

**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively. For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state. The blue line illustrates the mean of the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**eFigures 1 and 2**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 3**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 5**). For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease has been increasing over time for men in most states, with DC as a notable exception (**eFigure 5**). Recently, there has been a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with other causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**eFigure 7**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**eFigure 8**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Communicable disease is the rarest cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (Figures 15 and 16). We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969. In 2013, this proportion reached 53%. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from in infant mortality.

@@

Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20^th^ Century. [add: Lemann, 1991] Mortality trends for women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi.  On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy [@dwyer2016us]. On the other hand, greater racial equality can be achieved by falling black rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends. [add: Case and Deaton, 2017]

Consistent with what has been reported at the national level, we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time [@harper2012trends]. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time.  States pursue varying approaches to an array of relevant programs and laws including tobacco control [@cotti2016effects], Medicare coverage [@grogan2017racial], hand-gun violence [@humphreys2017evaluating] and nutrition [@taber2011state]. These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially effect blacks [@adimora2014policies]. The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others.  Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations, which is an advantage of using states as the units of analysis as opposed to larger regions or smaller counties, neither of which generate policy variations.[add: Strumpf FE Chapter in MSE]

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group [@case2015rising; @minton2016two]. The Arriaga decomposition method accounts for the fact that deaths occurring in one age or cause stratum are reflected in the absence of that death in another stratum. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. This may be especially consequential for blacks in states like Florida or New York that received many black Hispanic immigrants from Cuba, Puerto Rico and the Dominican Republic, potentially changing the composition of the state’s black population to some extent during the study period [@laveist2011black].  But it also has implications for the white population, especially in states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [@ruiz2013hispanic]. We were unable to adjust for Hispanic ethnicity in this study because the indicator for ethnicity on the death certificate has been noted to have varying accuracy across states and during the earlier time period [ref]. In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [@harper2014trends], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings overall. As the US population becomes increasingly Hispanic over time, however, this distinction grows more salient [@brazil2015spatial].

Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [@preston2014anatomy]. As we are examining the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC [add: US Census bureau, 2015].

@@

Racial differences have been a fixture of American life for centuries, but their continued evolution is a topic of great relevance for public policy and social justice.  The careful surveillance of these patterns is therefore a crucial public health activity, and this paper presents a unique perspective on state-level variation over nearly half a century. This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death. Differences in smoking policy, Medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.
